### 2022-04-05 chapters/conclusions/conclusions.tex
additional modifs post physics

\modif{Answering the aforementioned research questions required the development of new modeling tools. As this thesis relied heavily on Lagrangian transport modeling, several features were added to SLIM's particle tracker. Although the transport of the particles in itself was already implemented in the model, we developed features to characterize the state of the particles and compute connectivity indicators. The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decay rate of diseased agent mass or the diameter of sediment particles. Post-processing features were developed to compute the concentration or the deposition of particles. We also developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the connectivity networks. Additionally, we developed an epidemiological model using the computed connectivity matrices to simulated the spread of the stony coral tissue loss disease throughout Florida's Coral Reef. Finally, the particle tracker was made more flexible by developing new readers for different ocean model outputs such as HYCOM or the Nucleus for European Modelling of the Ocean (NEMO; \citealp{madec2017nemo}).}

\modif{Additionally, coupling SLIM with SWAN required some modifications of SLIM. First, we implemented a new parameterization of the wind drag coefficient to account for the formation of spume at the ocean-atmosphere interface in heavy wind conditions. An additional term was also added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane. Using SWAN also required the implementation of tools to process wave spectrum and parameters in SLIM. The implementation of the coupling between the two models followed the same approach as the ADCIRC+SWAN model. It required the development of tools to convert SLIM outputs and forcing into SWAN format, and to allow SLIM to read SWAN outputs. Compared to ADCIRC+SWAN, an additional challenge that we had to overcome was the fact that SLIM is based on discontinous Galerkin finite element models while SWAN requires continuous fields on the mesh. This was especially critical when defining the communication between the nodes of the partitions of SWAN based on the partitioning of the mesh used in SLIM.}

@@

\modif{\subsection*{Better understanding the impact of hurricanes on corals}}
\modif{In this thesis, the impact of hurricanes on coral reef ecosystems was mostly evaluated by modeling wave-induced advective transport processes. However, when studying the propagation of drifting particles, assessing the impact of hurricanes on the dispersion of the particle cloud is also of interest. This could be performed by computing the distance between the positions of the particles of the cloud and its center of mass or by evaluating the evolution of the concentration of particles in the cloud. Furthermore, wave-induced transport is not the only way tropical cyclones affect coral reef biota. First, they can cause a wholesale destruction of the reefs as they pass over them. However, this destruction can have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}. Furthermore, hurricanes increase water turbidity and induce coral burial by resuspending sediments. These impacts could be assessed by combining the developed coupled wave-current model with our sediment transport model. Moreover, hurricane interact with the Florida Current by disrupting its vertical structure and reducing its transport. These structural perturbations might have consequences on eddy activity and baroclinic instabilities near the Florida Keys. Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida although baroclinic models might be required. Finally, by modifying the ocean circulation on their passage, hurricanes might influence the connectivity of the Florida Reef Tract. The coupled SLIM+SWAN model could thus be used to evaluate the impact of Hurricane Irma on coral larval connectivity as a continuation of the work initiated in chapter \ref{chap:irma}. Furthermore, the hurricane-induced transport might have impacted the propagation of SCTLD through Florida's Coral Reef. Hurricane Irma could therefore have acted as a "super spreader" of disease agents or as a barrier to the propagation of the outbreak. This effect could be investigated by combining SLIM+SWAN with the developed epidemiological model.} \newline

>>>>

\modif{Answering the aforementioned research questions required the development of new modeling tools. As this thesis relied heavily on Lagrangian transport modeling, several features were added to SLIM's particle tracker. Although the transport of the particles in itself was already implemented in the model, we developed new model features to characterize the state of the particles and compute new connectivity indicators. The particle state can be used to represent biological traits such as the competency of coral larvae or abiotic characteristics such as the decay rate of diseased agent mass or the diameter of sediment particles. Post-processing features were developed to compute the concentration of suspended and deposited particles. We also developed tools to derive connectivity indicators such as the shortest paths between a pair sub-reef polygons in the connectivity networks. Additionally, we developed an epidemiological model using the computed connectivity matrices to simulated the spread of the stony coral tissue loss disease throughout Florida's Coral Reef. Finally, the particle tracker was made more flexible by developing new readers for different ocean model outputs such as HYCOM or the Nucleus for European Modelling of the Ocean (NEMO; \citealp{madec2017nemo}).}

\modif{Additionally, coupling SLIM with SWAN required some modifications of SLIM. First, we implemented a new parameterization of the wind drag coefficient to account for the formation of spume at the ocean-atmosphere interface in heavy wind conditions. An additional term was also added to the model's equations to account for the gradient of atmospheric pressure generated at the center of the hurricane. Using SWAN also required the implementation of tools to process the wave spectrum and the associated parameters in SLIM. The implementation of the coupling between the two models followed the same approach as the ADCIRC+SWAN model. It required the development of tools to convert SLIM outputs and forcing into SWAN format, and to allow SLIM to read SWAN outputs. Compared to ADCIRC+SWAN, an additional challenge that we had to overcome was the fact that SLIM is based on discontinous Galerkin finite element models while SWAN requires continuous fields on the mesh. This was especially critical when defining the communication between the nodes of the partitions of SWAN based on the partitioning of the mesh used in SLIM.}

@@

\modif{\subsection*{Better understanding the impact of hurricanes on corals}}
\modif{In this thesis, the impact of hurricanes on coral reef ecosystems was mostly evaluated by modeling wave-induced advective transport processes. However, when studying the propagation of drifting particles, assessing the impact of hurricanes on the diffusion (\ie~spread) of the particle cloud is also of interest. This could be performed by computing the distance between the positions of the particles of the cloud and its center of mass or by evaluating the evolution of the concentration of particles in the cloud. Furthermore, wave-induced transport is not the only way tropical cyclones affect coral reef biota. First, they can cause a wholesale destruction of the reefs as they pass over them. However, this destruction can sometimes have positive consequences by facilitating asexual recruitment and cleaning algal overgrowth on the reefs \citep{banks2008reef}. Moreover, water mixing and upwelling by hurricane can reduce water temperature and alleviate the thermal stress experienced by corals \citep{wilkinson2008status}. Furthermore, hurricanes increase water turbidity and induce coral burial by resuspending sediments. These impacts could be assessed by combining the developed coupled wave-current model with our sediment transport model. Moreover, hurricane interact with the Florida Current by disrupting its vertical structure and reducing its transport \citep{oey2007hurricane}. These structural perturbations might have consequences on eddy activity near the Florida Keys and therefore impact nutrient concentration and larval recruitment in the region \citep{lee1992influence}. Our study could therefore be extended by evaluating the impact of Hurricane Irma on the ocean state and the broader circulation patterns in Florida although baroclinic models might be required. Finally, by modifying the ocean circulation on their passage, hurricanes might influence the connectivity of the Florida Reef Tract. The coupled SLIM+SWAN model could thus be used to evaluate the impact of Hurricane Irma on coral larval connectivity as a continuation of the work initiated in chapter \ref{chap:irma}. Furthermore, the hurricane-induced transport might have impacted the propagation of SCTLD through Florida's Coral Reef. Hurricane Irma could therefore have acted as a "super spreader" of disease agents or as a barrier to the propagation of the outbreak. This effect could be investigated by combining SLIM+SWAN with the developed epidemiological model.} \newline

### 2022-04-05 chapters/intro/intro.tex
additional modifs post physics

\begin{figure}
    \centering
    \includegraphics[width=.8\textwidth]{chapters/intro/figures/anthropocene.png}
    \caption{Goods and services of coral reefs and their linkage with human activities. Source: \cite{hughes2017coral}}
    \label{intro:corals}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/intro/figures/cover.jpeg}
    \caption{Decline of the absolute coral cover in the Caribbean from $\sim$50\% to $\sim$10\% between 1977 and 2001. Source: \cite{gardner2003long}}
    \label{intro:cover}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/frt.png}
    \caption{Map of Florida Reef Tract with major features and divisions. Source: \cite{kupfner2019untapped}}
    \label{intro:frt}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
    \caption{Observed propagation of the \modif{stony coral tissue loss disease} throughout the \modif{Florida Reef Tract} between 2014 and 2021}
    \label{intro:propagation}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
    \caption{Total number of major hurricanes strikes in Florida between 1900 and 2010. Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
    \label{inro:landfall}
\end{figure}

@@

In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}). This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics of the whole Great Barrier Reef [ADD OTHER STUFF] \citep{lambrechts2008multi}. \modif{The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities. This model is therefore well suited for shallow areas where the water column is well mixed. Furthermore, assuming homogeneous vertical profiles reduces the computational cost and allows a finer spatial resolution of the model. However, in deeper areas or in regions where there is an inflow of freshwater, 3D baroclinic models are required to capture the variations of density in the water column due to temperature and salinity gradients. These differences in density can cause baroclinic instabilities that generate mesoscale eddies. Nonetheless, 3D baroclinic models are computationally more expensive, which prohibits the use of a reef-scale resolution on the whole Florida Reef Tract. As most Florida reefs are located in areas shallower than 30 meters (Fig. \ref{intro:bathy}), we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys. Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT with SLIM, \ie~the exchanges of larvae between reefs. [TODO: move somewhere else]The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements, as illustrated in appendix \ref{onset:appendice} and chapter \ref{chap:irma}}.

@@

\modif{Transport processes are at the core of this thesis. They can be estimated using two general methods: Eulerian tracers and Lagrangian particles. Eulerian methods were the earliest methods applied for transport modeling and are still widely used today. They offer the advantages of a fixed grid by solving the transport equations directly on the native grid of the model and are easy to implement \citep{spivakovskaya2007lagrangian}. The Lagrangian methods, on the other hand, follow virtual particles whose movements are modeled with a stochastic differential equation consistent with the advection-diffusion equation \citep{spivakovskaya2007lagrangian,van2018lagrangian}. One key difference between the two  methods is their computational cost. For each time step, the movements of $N$ virtual particles require $N$ sets of computation, while the advection-diffusion of a tracer takes as many sets of computation as there are discrete grid cells in the model. Additionally, Lagrangian methods allow the storage of the entire trajectory history of the virtual particles. This enables the \textit{a posteriori} computation of connectivity between different regions of the ocean \citep{thomas2014numerical,frys2020fine}. Finally, Lagrangian particles can be advected backward in time in offline mode, when velocity fields are stored. This reverse-time analysis gives a spatial probability distribution of places from where it is likely that particles originated to reach a given location \citep{spivakovskaya2005simulation}. In this thesis, transport processes were therefore studied using Lagrangian particles advected by SLIM currents. To answer our different research questions, we adapted SLIM transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.}

@@

%\modif{\section{Modeling framework and assumptions}}

%\modif{Wave-Current coupling. Radiation stress vs vector force}

%\modif{Presentation of the epidemiological model ?}

%\modif{Hypotheses of the model/study}\newline

>>>>

\begin{figure}
    \centering
    \includegraphics[width=.8\textwidth]{chapters/intro/figures/anthropocene.png}
    \caption{Goods and services of coral reefs and their linkage with human activities. \modif{Despite covering less than 0.5\% of the ocean floor, coral reefs are home for almost of third of the marine biodiversity and are a source of food and livelihood to hundreds of millions of people.} Source: \cite{hughes2017coral}}
    \label{intro:corals}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/intro/figures/cover.jpeg}
    \caption{Decline of the absolute coral cover in the Caribbean from $\sim$50\% to $\sim$10\% between 1977 and 2001. \modif{This illustrates the worldwide decline in live coral cover caused by both global and local stressors} .Source: \cite{gardner2003long}}
    \label{intro:cover}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/frt.png}
    \caption{Map of Florida Reef Tract with major features and divisions. \modif{The northern half-section of the Florida Reef Tract consists of relic, Holocene framework reefs with fairly low living hard coral cover. The southern hal-section is a chain of limestone islands (the Keys), that extend from the southern tip of Florida to the Dry Tortugas}. Source: \cite{kupfner2019untapped}}
    \label{intro:frt}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
    \caption{Observed propagation of the \modif{stony coral tissue loss disease} throughout the \modif{Florida Reef Tract} between 2014 and 2021. \modif{The disease was first observed off the coast of Miami in 2014, and then propagated through Florida's Coral Reef both northward and southward. The propagation of the disease appeared to slow down in 2020 between the Lower Keys and the Dry Tortugas (DRTO), before it finally reached this area in May 2021.}}
    \label{intro:propagation}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
    \caption{Total number of major hurricanes strikes in Florida between 1900 and 2010. \modif{Southern Florida is a prime landfall for tropical cyclones and has been directly hit by 27 of them during the past century.} Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
    \label{inro:landfall}
\end{figure}

@@

\modif{The motion of geophysical fluids such as seawater is characterized by the effects of ambient rotation and stratification \citep{cushman2011introduction}. The ambient rotation is induced by the Earth's spin and generates the Coriolis forces. Stratification arises from the presence of waters with different temperature and salinity, and therefore different density. Under the action of gravitational forces, heavier waters lower while lighter waters rise, which caused the water column to tend towards a stable stratified equilibrium \citep{cushman2011introduction}. However, this equilibrium is disrupted in the presence of discontinuities in horizontal velocities induced by wind and water motion \citep{miles1959generation}. These baroclinic instabilities generate internal waves, water mixing, and eddies \citep{cushman2011introduction}. Eddy activity is especially strong in boundary currents such as the Gulf Stream, where strong horizontal shear occur and where shelf water and deep water masses meet on the shelf break \citep{kourafalou2012florida}. These baroclinic instabilities were shown to have a significant impact on water composition and larval recruitment in the Florida Keys \citep{lee1992influence}. Capturing such processes require baroclinic models that represent the variations of water density along the water column. However, in regions where the water column is sufficiently homogeneous and shallow that vertical contrasts are negligible, ocean circulation is accurately approximated by barotropic models. Such models use a depth-averaged approach by assuming an homogeneous water column with constant density and are used to model estuaries, coastal regions and shelf seas \citep{tan1992shallow}. Both the baroclinic and the barotropic models usually rely on the hydrostatic equilibrium, \textit{i.e.} vertical accelerations are neglected by assuming that gravity balances the vertical component of the pressure gradient force. Such assumption is only valid when the horizontal scale of the flow is larger than its vertical scale \citep{cushman2011introduction}.}

In this thesis, we used the 2D barotropic version of the unstructured-mesh Second-generation Louvain-la-Neuve Ice-ocean Model (\href{https://www.slim-ocean.be/}{SLIM}). This model relies on Discontinuous Galerkin finite element methods \citep{aizinger2002discontinuous} and has already been successfully used to model the hydrodynamics \modif{of coastal regions and estuaries such as the whole Great Barrier Reef \citep{lambrechts2008multi}, the Florida Reef Tract \citep{frys2020fine}, the Scheldt estuary \citep{de2012water}, or the Mahakam delta \citep{sassi2011tidal}}. The 2D barotropic version of SLIM solves the shallow water equations, which assume a constant water density and compute depth-averaged current velocities. This model is therefore well suited for shallow areas where the water column is well mixed. Furthermore, assuming homogeneous vertical profiles reduces the computational cost and allows a finer spatial resolution of the model. However, in deeper areas or in regions where there is an inflow of freshwater, 3D baroclinic models are required to capture the effects of density discontinuities in the water column due to temperature and salinity gradients. Nonetheless, 3D baroclinic models are computationally more expensive, which currently prohibits the use of a reef-scale resolution on the whole Florida Reef Tract. As most Florida reefs are located in areas shallower than 30 meters (Fig. \ref{intro:bathy}), we coupled the 2D version of SLIM with GOM-HYCOM to indirectly capture the impact of baroclinic instabilities on the connectivity of the Florida Keys. Using this approach, \cite{frys2020fine} successfully modeled coral larval connectivity in the FRT with SLIM, \ie~the exchanges of larvae between reefs. \modif{The results of the model used in this thesis were validated against observations and showed good agreement with tide gauge and ADCP measurements \citep{dobbelaere2022connecting,dobbelaere2022impacts}}.

@@

\modif{Transport processes are at the core of this thesis. They can be estimated using two general methods: Eulerian tracers and Lagrangian particles. Eulerian methods were the earliest methods applied for transport modeling and are still widely used today. They offer the advantages of a fixed grid by solving the transport equations directly on the native grid of the model and are easy to implement \citep{spivakovskaya2007lagrangian}. The Lagrangian methods, on the other hand, follow virtual particles whose movements are modeled with a stochastic differential equation consistent with the advection-diffusion equation \citep{spivakovskaya2007lagrangian,van2018lagrangian}. One key difference between the two  methods is their computational cost. For each time step, the number of operations required to compute the movements of virtual particles is proportional to their number, while the computational cost of the advection-diffusion of a tracer is proportional to the number of discrete grid cells in the model. Additionally, Lagrangian methods allow the storage of the entire trajectory history of the virtual particles. This enables the \textit{a posteriori} computation of connectivity between different regions of the ocean \citep{thomas2014numerical,frys2020fine}. Finally, Lagrangian particles can be advected backward in time in offline mode, using an offline velocity field. This reverse-time analysis gives a spatial probability distribution of places from where it is likely that particles originated to reach a given location \citep{spivakovskaya2005simulation}. In this thesis, transport processes were therefore studied using Lagrangian particles advected by SLIM currents. To answer our different research questions, we adapted SLIM transport model to evaluate the dispersal of the causative agent of the SCTLD and sediments particles.}

@@

\modif{As this thesis is mostly based on modeling tools, it relies on the main assumption that a mechanistic model can accurately capture transport processes in Florida's Coral Reef. Models are simplifications of the reality and hypotheses must therefore be made on the processes driving the evolution of the system under study. Such hypotheses were for instance made by favoring a fine resolution barotropic model over a coarser baroclinic model of the Florida Reef Tract. Furthermore, we assumed that if the modeled waves and hydrodynamics accurately reproduced local observations from buoys, tide gauges and ADCP, then the modeled transport that they drive would be reasonably accurate as well. According to the popular saying, "all models are wrong but some are useful". In the present study, model results are confronted against field observations and experimental results to identify the potential drivers of the dispersion of stressors in Florida's Coral Reef.}

### 2022-04-05 chapters/irma/article.tex
additional modifs post physics

\begin{figure}
    \centering
    \includegraphics[width=.99\textwidth]{chapters/irma/figures/coupling_v2.png}
    \caption{Schematic illustration of the coupled SLIM+SWAN model.}
    \label{fig:coupling}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/irma/figures/figure_initial_positions.png}
    \caption{Release regions of the passive particles on the inner and outer shelves on Sept 7 at 0000 UTC (red and blue clouds) obtained by backtracking particles released in the red and blue circular areas during the passage of Irma.}
    \label{fig:init}
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=.99\textwidth]{chapters/irma/figures/coupling_v2.png}
    \caption{Schematic illustration of the coupled SLIM+SWAN model. \modif{Both models run on the same computational cores and mesh. SLIM is run first and passes the wind velocity $U_{10}$, water level $\eta$, depth-averaged current $u$ and bottom roughness length $z_0$ to SWAN. SWAN then uses these quantities to compute the wave radiation stress gradient $\text{\boldmath$\tau$}_\text{wave}$, that is then passed
    to SLIM as the force exerted by waves on currents. SLIM then uses this quantity to update the value of the surface stress $\text{\boldmath$\tau$}_s$, computed as the combination of wind and wave stresses $\text{\boldmath$\tau$}_\text{wind} + \text{\boldmath$\tau$}_\text{wave}$}}
    \label{fig:coupling}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.95\textwidth]{chapters/irma/figures/figure_initial_positions.png}
    \caption{Release regions of the passive particles on the inner and outer shelves on Sept 7 at 0000 UTC (red and blue clouds) obtained by backtracking particles released in the red and blue circular areas during the passage of Irma. \modif{These clouds give a probabilistic picture of the positions from which particles should have been released to intersect the path of the center of the hurricane}}
    \label{fig:init}
\end{figure}
